Taking Dyads Seriously

### 2018-02-12 paper/1_intro.tex
intro edits

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, \textit{International Studies Quarterly} published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012,franzese:hayes:2007,king:roberts:2014}.

In this article, we present the Additive and Multiplicative Effects (AME) model, a Bayesian approach for directly modeling relational data to reflect better the interdependencies underlying the data generating process of dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still essential to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict against a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third order dependencies capture relationships of transitivity, balance, and clusterability between different dyads--for example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003 if we understand that the United States invaded Iraq in 2003 and that Poland often participates in US-led coalitions. The multiplicative effects capture these sorts of dependencies, especially those that result because the specified model has not accounted for some latent set of shared attributes possessed by actors that affect their probability of interacting with one another.

@@

The framework that we present advances statistical analysis of dyadic data by accounting for observational dependence while easily testing the substantive effect of variables of interest. Thus, the AME approach can be used by scholars in the field to not only generate substantive insights, but to sufficiently model the data generating process behind political events of interest. Most importantly, it concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.

>>>>

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, \textit{International Studies Quarterly} published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to test hypothesized relationships.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012,franzese:hayes:2007,king:roberts:2014}.

In this article, we present the Additive and Multiplicative Effects (AME) model, a Bayesian approach for directly modeling relational data to better reflect the interdependencies underlying the data generating process of dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still essential to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict against a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third order dependencies capture relationships of transitivity, balance, and clusterability between different dyads--for example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003 if we understand that the United States invaded Iraq in 2003 and that Poland often participates in US-led coalitions. The multiplicative effects capture these sorts of dependencies, especially those that result because the specified model has not accounted for some latent set of shared attributes possessed by actors that affect their probability of interacting with one another.

@@

The framework that we present advances statistical analysis of dyadic data by accounting for observational dependence while enabling scholars to still focus on testing the substantive effect of variables of interest. Thus, the AME approach can be used by scholars in the field to continue to generate substantive insights, while having the benefit of being able to better account for the data generating process behind political events of interest. Most importantly, it concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.

### 2018-02-12 paper/4_app.tex
intro edits

\documentclass{article}[12pt]
\usepackage{setspace}
\usepackage{listings}
\doublespacing
\usepackage[letterpaper, portrait, margin=1in]{geometry}
\setlength{\parskip}{.5em}
\setlength{\parindent}{0cm}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{framed}
\usepackage[sort,round]{natbib}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{hyperref}

\begin{document}

@@

%% What is the point:

@@

% SM: maybe too on the nose (Whether putative control variables, for
% example, are precisely estimated is less important.) 
%% MJF: 2/9-- took this part out, because I don't think that it's necessar.
% Typically, scholars are concerned with conducting hypothesis testing on a small set of independent variables that they have assembled in a complicated empirical model. For example, in the study by Reiter and Stam (2003) the goal is to determine whether in mixed dyads---consisting of a democratic country and country ruled by a personalist dictatorship---it is the democratic country or the personalist country that is associated with a greater likelihood of a dispute dyad being in a dispute.  In a regression including fourteen covariates, they focus on whether the democratic nature of the initiating country is a statistically significant covariate.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our re-examination of Reiter and Stam, we focus on whether the model replicates when re-estimated within the AME framewo
%rk.
Each of the five studies has a crucial finding that we hone in on to
further draw into focus the analytical power of the AME estimation
procedure.  In Table~\ref{tab:modelFindingSumm}, we present the
overall results; the term \textit{Remains after accounting for
  dependencies?} indicates only that the sign and/or significance of
the putatively crucial finding in the original study is not found to
hold in the AME estimation.\footnote{Full tabular results for each of
  the original and reestimated models are presented in the Appendix.}

@@

An important takeaway here is that %as a result of the incompatibility between GLM models and the underlying processes that shape international events, 
many scholars are forced to make knowledge claims based on the statistical significance of a small set of covariates, or
the differences between these covariates.  These differences may
change dramatically when interdependencies are taken into account
directly, as they do in the case of this study. %MJF: Keeping this here for now, but it reads more like a conclusion
This outcome follows from AME's ability to precisely capture higher-order
interdependencies, whereas GLM approaches implicitly ignore dyadic
interdependencies. As this is a widely-known limitation of GLM
approaches, scholars often attempt to account for clustering of
observations by including additional variables and adjusting the
standard errors of the resulting estimates. At best, this method
introduces noise and imprecision into results, and at worst can
produce misleading outcomes.   Beyond just comparing parameter
estimates, we examine how well each approach can represent the data
generating process using an out-of-sample cross validation
strategy. Specifically, for each study, we randomly divide the data
into $k=30$ sets, letting $s_{ij,t}$ be the set to which pair $ij,t$
is assigned.

@@

\end{document}

>>>>

@@

@@

% SM: maybe too on the nose (Whether putative control variables, for
% example, are precisely estimated is less important.) 
%% MJF: 2/9-- took this part out, because I don't think that it's necessar.
% Typically, scholars are concerned with conducting hypothesis testing on a small set of independent variables that they have assembled in a complicated empirical model. For example, in the study by Reiter and Stam (2003) the goal is to determine whether in mixed dyads---consisting of a democratic country and country ruled by a personalist dictatorship---it is the democratic country or the personalist country that is associated with a greater likelihood of a dispute dyad being in a dispute.  In a regression including fourteen covariates, they focus on whether the democratic nature of the initiating country is a statistically significant covariate.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our re-examination of Reiter and Stam, we focus on whether the model replicates when re-estimated within the AME framewo
%rk.

Each of the five studies has a crucial finding that we hone in on to further draw into focus the analytical power of the AME estimation procedure.  In Table~\ref{tab:modelFindingSumm}, we present the overall results; the term \textit{Remains after accounting for dependencies?} indicates only that the sign and/or significance of the putatively crucial finding in the original study is not found to hold in the AME estimation.\footnote{Full tabular results for each of the original and reestimated models are presented in the Appendix.}

@@

% as a result of the incompatibility between GLM models and the underlying processes that shape international events, 
%MJF: Keeping this here for now, but it reads more like a conclusion

An important takeaway here is that many scholars are forced to make knowledge claims based on the statistical significance of a small set of covariates, or the differences between these covariates.  These differences may change dramatically when interdependencies are taken into account directly, as they do in the case of this study. This outcome follows from AME's ability to precisely capture higher-order interdependencies, whereas GLM approaches implicitly ignore dyadic interdependencies. As this is a widely-known limitation of GLM approaches, scholars often attempt to account for clustering of observations by including additional variables and adjusting the standard errors of the resulting estimates. At best, this method introduces noise and imprecision into results, and at worst can produce misleading outcomes.   Beyond just comparing parameter estimates, we examine how well each approach can represent the data generating process using an out-of-sample cross validation strategy. Specifically, for each study, we randomly divide the data into $k=30$ sets, letting $s_{ij,t}$ be the set to which pair $ij,t$ is assigned.

@@
